---
title: "Aggregating data"
author: "Carole Voulgaris"
date: "9/3/2020"
output: 
  html_document:
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

For this tutorial, I'll be demonstrating how to aggregate person-level data to the household level.

```{r load libraries, message=FALSE}
library(tidycensus)
library(tidyverse)
```

I'll start by downloading a dataset from the ACS 1-year sample from 2018 for Rhode Island. I'm limiting my sample to people who do not live in group quarters (indicated by those with a "b" value for the TEN variable).

I'm including two variables:

1. Income (PINCP) (person-level variable)
2. Number of bedrooms (BDSP) (household-level variable)

Since one is a person-level variable, the resulting dataset will be a person-level dataset with a row for each person. All people in the same household will have the same value for the household level variable(s).

``` {r load person-level data, results = 'hide'}
person_data <- get_pums(variables = c("PINCP", "BDSP", "TEN"),
                        state = "RI", 
                        year = 2018, 
                        survey = "acs1") %>%
  filter(TEN != "b")
```
Here's what the data looks like:

```{r show household data}

person_data
  
```

In addition to the variables I requested, I also get the following variables:

1. **SERIALNO** This is an ID number that identifies each unique household.
2. **SPORDER** This is an ID number that identifies each individual within the household. There are four people in the first household listed above, then the next households each have one person in them, and those are followed by a household with three people in it.
3. **WGTP** and **PWGTP** These are weighting variables that you would need to aggregate your data get estimates for an entire geography (e.g. a county or a census tract), which is beyond to scope of this course (to get those estimates, just use `get_acs()`).

To aggregate to the household level, I can use `group_by()` to group together individuals from the same household and `summarize` to calculate summary values within those groups. 
* `n()` returns the number of rows in each group, so that will give me the number of people in each household.
* `sum()` returns the total of the values within each group, so by adding up the income of each person, I can get the total household income.
* `mean()` returns the average within each group. Since BDSP is a household-level variable, it has the same value for each person in the household, and the average will be the same as that value.

Note that some people in `people_data` have negative values for income. This is a code to indicate that income data was not collected for that person (for example, because they are children), so before adding up the incomes, we should set those values to zero.

```{r}
hh_data <- person_data %>%
  mutate(PINCP = ifelse(PINCP < 0, 0, PINCP)) %>%
  group_by(SERIALNO) %>%
  summarize(n_people = n(),
            tot_income = sum(PINCP),
            n_bedrooms = mean(BDSP))

hh_data
```

Compare hh_data to people_data to confirm that the values are what you would expect them to be.