---
title: "Week 2: Distributions and proportions (calculations only)"
author: "Carole Voulgaris"
date: "9/10/2020"
output: 
  html_document:
    theme: cosmo
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

For this assignment, you'll be using the following libraries, so make sure they're installed ahead of time.

```{r load libraries}
library(tidyverse)
library(ggplot2)
```

Let's start by loading the person-level dataset we created last week.

`read_csv()` loads the data into our environment.

```{r load data, message=FALSE}
person_data <- read_csv("people.csv")
```

## Distribution of a continuous variable
Let's start by looking at the distribution of the amount of time (in minutes) people spend traveling to work. 

### Summary statistics
The `summary()` function can quickly give us a quick summary 

```{r quick summary statistics}
JWMNP_summary <- summary(person_data$JWMNP)

JWMNP_summary
```

I can access the individual parts of the JWMNP_summary object as well. For example, maybe I'm only interested in the mean.

```{r}
JWMNP_summary["Mean"]
```

Or maybe I'm only interested in the value of the third quartile.

```{r}
JWMNP_summary["3rd Qu."]
```

On the other hand, if I'm only interested in the mean, it's even easier to just use the `mean()` function.

```{r}
mean(person_data$JWMNP)
```

Likewise, if I only want the median, I can use the `median()` function

```{r}
median(person_data$JWMNP)
```

If I only want the quartile values (including the minimum and maximum), I can use the `quantile` function.

```{r}
quantile(person_data$JWMNP)
```

And if I only want the 25th percentile, I can grab just that piece.

```{r}
quantile(person_data$JWMNP)["25%"]
```

I can use the function `sd()` to get the sample standard deviation (which isn't included in the output of the `summary()` function).

```{r}
sd(person_data$JWMNP)
```

To get a 95-percent confidence interval for the population mean, I'll use the t.test() function to do a one-sample t-test.

```{r}
conf_int <- t.test(person_data$JWMNP)

conf_int
```

I can grab just the center of that confidence interval, which will be the same as the sample mean.

```{r}
conf_int$estimate
```

And I can also get just the lower and upper limits of the confidence interval. Here's the lower limit:

```{r}
conf_int$conf.int[1]
```

And here's the upper limit:

```{r}
conf_int$conf.int[2]
```

### Histogram
All of the above numbers are useful for describing the distribution, and it also helps to visualize it with a histogram. `hist()` will generate a quick histogram for you.

```{r}
hist(person_data$JWMNP)
```

But I prefer to use `geom_histogram()` with `ggplot()` to take full advantage of the features of the `ggplot2` library.

```{r}
ggplot(person_data, aes(x = JWMNP)) +
  geom_histogram()
```

The distribution is skewed a little to the right. It actually looks like it might be closer to a Poisson distribution or negative binomial distribution than to a normal distribution. You might notice that there's a little bump in the number of people who report having a 60-minute journey to work. This is not unusual for self-reported time use. It's likely that people with 55-minute and 65-minute commutes are rounding to 60 minutes.

## Proportions of a categorical variable

We can use `unique()` to list all the possible values of a categorical vairable.

```{r}
unique(person_data$TEN_label)
```

It looks like there are five different tenure categories in my sample. What proportion of the sample is in each one?

I can use `table()` to get the number of observations in each category.

```{r}
table(person_data$TEN_label)
```

And I can convert those numbers to proportions by dividing by the total number of observations.

```{r}
table(person_data$TEN_label) / sum(table(person_data$TEN_label))
```

I can also calculate each of those proportions individually as the average of a binary (TRUE/FALSE) variable. For example, the proportion of the sample that live in homes they own free and clear:

```{r}
mean(person_data$TEN_label == "Owned free and clear")
```

I can use the `t.test()` function to get the 95-percent confidence interval for that proportion too.

```{r}
t.test(person_data$TEN_label == "Owned free and clear")
```
