---
title: "Cleaning Data from data.census.gov"
author: "Carole Voulgaris"
date: "9/7/2020"
output: 
  html_document:
    theme: cosmo
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Setup

Before I begin, I'll load the tidyverse package.

```{r, message = FALSE}
library(tidyverse)
```

## Loading data

I previously downloaded a zipped file named ACSDT5Y2018.B07009_2020-09-07T084016.zip from data.census.gov and saved it to my working directory. The zipped file contains three files:

1. The dataset (ACSDT5Y2018.B07009_data_with_overlays_2020-09-07T084012.csv)
2. A list of variable names and descriptions (ACSDT5Y2018.B07009_metadata_2020-09-07T084012.csv)
3. A description of the data source (ACSDT5Y2018.B07009_table_title_2020-09-07T084012.txt)

We want to extract the files and save the two csv files as dataframes.

``` {r load data, message = FALSE}
unzip("ACSDT5Y2018.B07009_2020-09-07T084016.zip")

variables <- read_csv("ACSDT5Y2018.B07009_metadata_2020-09-07T084012.csv")
data <- read_csv("ACSDT5Y2018.B07009_data_with_overlays_2020-09-07T084012.csv", skip = 1)
```

## Choosing variables

There are basically two variables in the table I've downloaded: Educational attainment and geographic mobility, so I probably need to download another table from data.census.gov to get the other variables I want for Assignment 1. For this example, let's say I want these two variables from the table:

1. Whether the majority of the tract population has at least a four-year college degree
2. The percentage of the population that has lived in the same house for the past year

``` {r derive variables}
data <- data %>%
  mutate(pct_not_moved = `Estimate!!Total!!Same house 1 year ago` /
                      `Estimate!!Total`,
         pct_college = 
           (`Estimate!!Total!!Bachelor's degree` +
             `Estimate!!Total!!Graduate or professional degree`) /
          `Estimate!!Total`) %>%
  mutate(majority_college = pct_college > 0.5) %>%
  mutate(geoid = substr(id, 10, 20)) %>%
  select(geoid, pct_not_moved, majority_college)
```

And here are the first few rows of the resulting dataset:

``` {r show data}
data
```