---
title: "Week 2: Distributions and proportions (including formatting)"
author: "Carole Voulgaris"
date: "9/10/2020"
output: 
  html_document:
    theme: cosmo
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

Once you've looked through the [calculations only](https://ses-5215-f2020.github.io/tutorials/Week2--calculatoins-only){target="_blank"} tutorial, you might look through this one for some examples that would be helpful for creating a formatted report with the results of your analysis.

I'll be using the following libraries, so make sure they're installed ahead of time.

```{r load libraries}
library(tidyverse)
library(ggplot2)
```

Let's start by loading the person-level dataset we created last week.

`read_csv()` loads the data into our environment and `attach()` basically activates it so that when we refer to one of the variables in that dataframe, R knows which dataframe to look for it in.

```{r load data}
person_data <- read_csv("people.csv")
attach(person_data)
```

## Distribution of a continuous variable
Let's start by looking at the distribution of the amount of time (in minutes) people spend traveling to work. The `summary()` function can quickly give us some summary 

```{r quick summary statistics}
summary(JWMNP)
```

I can also use the functions `mean()` and `quantile()` to get these pieces separately. We can also use `sd()` to get the sample standard deviation, which isn't included in the output of the `summary()` function.

```{r individual summary statistics}
paste("The sample mean for the travel time to work is",
      prettyNum(mean(JWMNP), digits = 0), "minutes.")
paste("The sample median for the travel time to work is",
      prettyNum(median(JWMNP), digits = 0), "minutes.")
paste("The sample standard deviation for the travel time to work is",
      prettyNum(sd(JWMNP), digits = 0), "minutes.")
paste("The sample interquartile range for the travel time to work is",
      quantile(JWMNP)["25%"], "to", quantile(JWMNP)["75%"], "minutes.")
```

All of the above are characteristics of the sample. We can use t.test() to get a 95-percent confidence interval for the population mean.

```{r population mean}
pop_mean <- t.test(JWMNP)
pop_mean
paste("The 95-percent confidence interval for the population mean is",
      prettyNum(pop_mean$conf.int[1], digits = 0), "to",
      prettyNum(pop_mean$conf.int[2], digits = 0), "minutes.")
```

We can get a quick look at that distribution by creating a histogram. We'll use the ggplot2 package for all our graphics. I'll save the graph to an object called `hist`, so I can add to it a little bit later.

```{r create histogram}
hist <- ggplot(person_data, aes(x = JWMNP)) +
  geom_histogram() 
hist
```

Let's pretty up that histogram a little, and display the summary statistics we just calculated right on the figure.

```{r better histogram}
hist + 
  theme_bw() +
  scale_x_continuous(name = "Travel time to work (minutes)") +
  scale_y_continuous(name = "Number of people in sample") +
  annotate("text", x = 50, y = 22500, size = 3, 
           label = paste("Population mean (95% confidence):",
                         prettyNum(pop_mean$conf.int[1], digits = 0),
                         "to",
                         prettyNum(pop_mean$conf.int[2], digits = 0),
                         "minutes.",
                         "\nSample mean:", 
                         prettyNum(mean(JWMNP), digits = 0), "minutes.",
                         "\nSample median:",
                         prettyNum(median(JWMNP), digits = 0), "minutes.",
                         "\nSample standard deviation:",
                         prettyNum(sd(JWMNP), digits = 0), "minutes.", 
                         "\nSample interquartile range:",
                         prettyNum(quantile(JWMNP)["25%"], digits = 0),
                         "to",
                         prettyNum(quantile(JWMNP)["75%"], digits = 0),
                         "minutes."),
           hjust = 0, vjust = 1) 
```

## Distribution of a categorical variable

The distribution of a categorical variable can be decribed in terms of the proportion of the sample or population that falls in each category. Let's take a look at the housing tenure variable (`TEN_label`).

First, we can get a list of all the possible values

```{r tenure categories}
tenures <- unique(TEN_label)
tenures
```

Now let's get the share of the sample in each category.

``` {r}
group_quarters <- t.test(TEN_label == tenures[1])
owned_free <- t.test(TEN_label == tenures[2])
mortgage <- t.test(TEN_label == tenures[3])
rented <-  t.test(TEN_label == tenures[4])
rent_free <-  t.test(TEN_label == tenures[5])
shares <- tibble(Tenure = c("Group Quarters", 
                            "Owned free and clear", 
                            "Owned with mortgage",
                            "Rented",
                            "Occupied rent-free"),
                 `Share` = c(group_quarters$estimate, 
                             owned_free$estimate,
                             mortgage$estimate,
                             rented$estimate,
                             rent_free$estimate),
                 Low = c(group_quarters$conf.int[1],
                         owned_free$conf.int[1],
                         mortgage$conf.int[1],
                         rented$conf.int[1],
                         rent_free$conf.int[1]),
                 High = c(group_quarters$conf.int[2],
                         owned_free$conf.int[2],
                         mortgage$conf.int[2],
                         rented$conf.int[2],
                         rent_free$conf.int[2]))
knitr::kable(shares, caption = "Proportions and 95-percent confidence intervals")
```

Now that we've calculated the confidence intervals for those proportions, and organized them into a dataframe, we can illustrate those results with a bar chart.

``` {r bar chart}
ggplot(shares, aes(x = Tenure, y = Share, ymin = Low, ymax = High)) +
  geom_bar(stat = "identity") +
  geom_errorbar(width = 0.5) +
  theme_bw() +
  scale_x_discrete(labels = c("Group Quarters", 
                            "Owned\nfree and clear", 
                            "Owned\nwith mortgage",
                            "Rented",
                            "Occupied\nrent-free")) +
  scale_y_continuous(name = "Share of population (error bars represent\n 95-percent confidence interval)",
                     breaks = c(0, 0.1, 0.2, 0.3, 0.4, 0.5),
                     labels = c("0", "10%", "20%", "30%", "40%", "50%"))
```
